Hierarchical time series forecasting in Emergency Medical Services

### 2022-11-17 manuscript/main.tex
first draft of results and conclusion

\begin{keyword}
    forecasting \sep healthcare \sep emergency services \sep forecast
reconciliation \sep hierarchical time series \sep ambulance demand \sep 
    attended incidents
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, sharp corners, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

\textbf{Poisson Regression using tscount:} We consider another Poisson
regression model that takes into account serial dependence in addition
to covariates used in the GLM model. To that end, we use
\texttt{tsglm()} function in \emph{tscount} package in R to model the
attended incidents. The logarithmic link function is used to ensure that
the parameter of Poisson distribution is always positive. This model can
be estimated via maximization of the likelihood function based on
Poisson mass function. The regression model captures the short range
serial dependence by including the three order autoregressive terms.

@@

There are many applications in the healthcare and in particular in EMS
where a collection of time series is available. These series are
generally hierarchically organized based on multiple levels such as
area/region, health board and/or are aggregated at different levels in
groups based on nature of demand, priority of demand or some other
attributes. While series could be strictly hierarchical or only grouped
bases on some attributes, in many situation a more complex structures
arise when attributes of interest are both nested and crossed, having
hierarchical and grouped structure. This is also the case as discussed
in Section~\ref{sec-data}

@@

\hypertarget{point-forecast}{%
\subsection{Point forecast}\label{point-forecast}}

\hypertarget{probabilistic-forecast}{%
\subsection{Probabilistic forecast}\label{probabilistic-forecast}}

@@

>>>>

\begin{keyword}
    forecasting \sep healthcare \sep emergency services \sep forecast
reconciliation \sep hierarchical time series \sep ambulance demand \sep 
    attended incidents
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

\textbf{Poisson Regression using tscount: } We consider another Poisson
regression model that takes into account serial dependence in addition
to covariates used in the GLM model. To that end, we use
\texttt{tsglm()} function in \emph{tscount} package in R
\citep{JSSv082i05} to model the attended incidents. The logarithmic link
function is used to ensure that the parameter of Poisson distribution is
always positive. This model can be estimated via maximization of the
likelihood function based on Poisson mass function. The regression model
captures the short range serial dependence by including the three order
autoregressive terms.

@@

There are many applications in the healthcare and in particular in EMS
where a collection of time series is available. These series are
generally hierarchically organized based on multiple levels such as
area/region, health board and/or are aggregated at different levels in
groups based on nature of demand, priority of demand or some other
attributes. While series could be strictly hierarchical or only grouped
bases on some attributes, in many situation a more complex structures
arise when attributes of interest are both nested and crossed, having
hierarchical and grouped structure. This is also the case as discussed
in Section~\ref{sec-data}.

@@

In this section, we compare the forecasting performance of Naive, ETS,
GLM, Tscount and the ensemble model using base forecast and Minimum
Trace (Mint) methods. We have also computed the forecast accuracy for
Ordinary Least Square (OLS) and Weighted Least Square (WLS) approaches.
However, they are not reported here because their accuracy is
outperformed by Mint. We should also note that forecasts, and
consequently their corresponding errors, are generated for the entire
hierarchy and they could be reported at any level, if required.

The overall forecasting performance is reported in
Table~\ref{tbl-result}, in which the average forecast accuracy per
model, method and the hierarchical level is presented. Reported forecast
accuracy are averaged across all forecast horizons, rolling origins and
series at each level. Table~\ref{tbl-result} presents both point and
probabilistic forecast accuracy at total, control area, health board and
bottom level series. Point forecast performance are reported using MASE
and RMSSE in Table~\ref{tbl-result-1} and Table~\ref{tbl-result-2},
respectively. Probabilistic forecast accuracy is reported using CRPS in
Table~\ref{tbl-result-3}. The bold entries in each table identify a
combination of method and model that performs best for the corresponding
level (i.e.~each column), based on the smallest values of accuracy
measures.

Table~\ref{tbl-result-1} shows that forecast reconciliation (i.e.~Mint)
improves forecast accuracy at the higher levels of the hierarchy
including total, control area and health board. However, it is does not
result in accuracy improvement at the bottom level series, fro which
base forecasts are more accurate. This might be due to the noisy
structure of time series at the bottom level and the fact that base
forecasts produced for the bottom of hierarchy are from well-specified
forecasting models and therefore does not gain benefit from forecast
reconciliation. Generally, if the bottom-level series are sufficiently
smooth, base forecast is expected to outperform hierarchical
reconciliation due to the better forecasts available at the level. It is
also clear from Table~\ref{tbl-result-1} that the ensemble method
improves forecast accuracy at total, control area and health board.
However, this does not remain valid for bottom series where the Poisson
regression by GLM outperforms other approaches.

Table~\ref{tbl-result-2} also reports point forecast accuracy but using
RMSSE. The outperformance of forecast reconciliation using Mint at
higher levels of hierarchy is also confirmed by RMSSE. However, for the
bottom series, we observe that the base forecasts generated by ETS are
the most accurate, followed by its reconciliations forecasts.

\begin{table}

\caption{\label{tbl-result}Average forecast perfomance calculated on
out-of-sample with time series cross validation applied to attended
incident data. The best approach is highlighted in
bold.}\begin{minipage}[t]{\linewidth}
\subcaption{\label{tbl-result-1}Point forecast accuracy using MASE }

\tabularnewline

\centering
\begin{tabular}[t]{llllll}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{4}{c}{MASE} \\
\cmidrule(l{3pt}r{3pt}){3-6}
method & model & Total & Control areas & Health boards & Bottom\\
\midrule
base & ensemble & 0.7543 & 0.8432 & 0.8648 & 1.004\\
base & ets & 0.9360 & 0.9133 & 0.8884 & 1.031\\
base & iglm & 0.8609 & 0.9083 & 0.9039 & \textbf{0.999}\\
base & naiveecdf & 1.1396 & 1.0633 & 1.0389 & 1.017\\
base & tscount & 0.8704 & 0.9109 & 0.9041 & 1.000\\
\addlinespace
mint & ensemble & \textbf{0.7332} & \textbf{0.8346} & \textbf{0.8621} & 2.251\\
mint & ets & 0.8593 & 0.9001 & 0.8997 & 1.258\\
mint & iglm & 0.8176 & 0.8805 & 0.8883 & 2.490\\
mint & naiveecdf & 1.1395 & 1.0631 & 1.0389 & 2.648\\
mint & tscount & 0.8218 & 0.8846 & 0.8913 & 2.516\\
\bottomrule
\end{tabular}

\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}
\subcaption{\label{tbl-result-2}Point forecast accuracy using RMSSE }

\tabularnewline

\centering
\begin{tabular}[t]{llllll}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{4}{c}{RMSSE} \\
\cmidrule(l{3pt}r{3pt}){3-6}
method & model & Total & Control areas & Health boards & Bottom\\
\midrule
base & ensemble & 0.7338 & 0.8335 & 0.8652 & 0.9885\\
base & ets & 0.9465 & 0.9126 & 0.8915 & \textbf{0.9795}\\
base & iglm & 0.8383 & 0.9073 & 0.9125 & 0.9958\\
base & naiveecdf & 1.0559 & 1.0213 & 1.0188 & 1.0081\\
base & tscount & 0.8446 & 0.9090 & 0.9114 & 1.0069\\
\addlinespace
mint & ensemble & \textbf{0.7153} & \textbf{0.8244} & \textbf{0.8621} & 1.2405\\
mint & ets & 0.8510 & 0.9002 & 0.9027 & 0.9879\\
mint & iglm & 0.8015 & 0.8814 & 0.8972 & 1.3379\\
mint & naiveecdf & 1.0558 & 1.0211 & 1.0188 & 1.4429\\
mint & tscount & 0.8020 & 0.8849 & 0.9003 & 1.3515\\
\bottomrule
\end{tabular}

\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}
\subcaption{\label{tbl-result-3}Probabilistic forecast accuracy using CRPS }

\tabularnewline

\centering
\begin{tabular}[t]{llllll}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{4}{c}{CRPS} \\
\cmidrule(l{3pt}r{3pt}){3-6}
method & model & Total & Control areas & Health boards & Bottom\\
\midrule
base & ensemble & 12.44 & 5.681 & 3.385 & 0.2421\\
base & ets & 13.96 & 5.974 & 3.430 & 0.2422\\
base & iglm & 14.49 & 6.061 & 3.488 & 0.2420\\
base & naiveecdf & 30.35 & 10.887 & 5.445 & 0.3000\\
base & tscount & 14.98 & 6.124 & 3.500 & 0.2429\\
\addlinespace
mint & ensemble & \textbf{12.09} & \textbf{ 5.664} & \textbf{3.383} & 0.2455\\
mint & ets & 13.28 & 5.884 & 3.472 & \textbf{0.2412}\\
mint & iglm & 13.35 & 5.812 & 3.394 & 0.2443\\
mint & naiveecdf & 30.36 & 10.904 & 5.446 & 0.3103\\
mint & tscount & 13.55 & 5.868 & 3.423 & 0.2483\\
\bottomrule
\end{tabular}

\end{minipage}%

\end{table}

Table~\ref{tbl-result-2} presents the accuracy of the forecast
distributions measures by CRPS, which considers both forecasting
reliability and interval sharpness. The smaller the value of CRPS, the
better the comprehensive performance. We observe that forecast
reconciliation results in forecast improvement, regardless of the
hierarchical level. Ensemble method is also more accurate fore higher
levels, but ETS performs better at the bottom level. While results show
that forecast reconciliation does not improve point forecast at the
bottom level, however Table~\ref{tbl-result-2} indicates that it
generates more accurate forecasts than base method. The outperformance
of probabilistic forecast reconciliation at the bottom level may
highlight an important fact. The reconciliation method may improve
forecast accuracy at certain number of quantiles at the tails of
forecast distribution, which are critical for managing risks.

Overall, our results indicate that forecast reconciliation using MinT
method provides reliable forecasts, and all methods that reconcile using
information at all levels of the forecast improve upon base
(unreconciled) forecasts, except the bottom level series. For bottom
series with less obvious patterns, base forecasts near center of
forecast distribution might be more accurate, while forecast
reconciliation using Mint could improve accuracy in the tails of of the
distribution.

In addition to the overall forecast accuracy presented in
Table~\ref{tbl-result}, we also report the point and probabilistic
forecast accuracy measures for each forecast horizon in
Figure~\ref{fig-accuracy}. The figure focuses on the hierarchical levels
important for decision making including total, control area and health
board, however this could be calculated for any level. We only
illustrate the results of the Mint method, given the its outperformance
described in Table~\ref{tbl-result}. For the illustration purpose, we
report the average weekly forecast accuracy instead of the daily
forecast horizon (i.e.~84 days including freezed planning (7 weeks = 42
days) plus future planing period (7 weeks = 42 days). Therefore, x-axis
shows horizons from week 1 (h= 1,\ldots,7) to week 12 (h= 78,\ldots,84).
The forecast horizon from week 7 to week 12 corresponds to the upcoming
planning horizon, which is used by planners and decision making. For the
point forecast, we can see that ensemble model performs much better than
others regardless of forecast horizons. For the probabilistic forecasts,
the distinction between forecasting models is less obvious. It is
important to highlight that, all forecasting models outperform the Naive
empirical distribution that is used as a Benchmark for both point and
probabilsitic forecasts.

\begin{figure}[H]

@@

Forecasting problems at Emergency Medical Services have often inherent
hierarchical and grouped structures. For example, looking at time series
of arrival calls in a clinical desk service, verified incidents, or
attended incidents in a country, they could be disaggregated by various
attributes of interest. Total demand in the country could be
disaggregated by region, then within each region by health board, within
each health board, by station/hospital, and so on down to the post code
area. Alternative structure may arise when attributes of interest are
crossed rather than nested. For example, the total demand could be
disaggrgated by priority (e.g Red, Amber, Green) or nature of incidents.
It is also natural to have a mixed structures, for example the total
demand could be disaggregated by priority and also by health board.

Despite the inherent hierarchical structure of the forecasting problem
in EMS, the common practice is to produce point forecast for each time
series independently. This practice may lead to the lack of coordination
and possibly undesirable and conflicting outcomes. Furthermore, due to
the asymmetric impact of resource allocation in this area, quantifying
forecast uncertainty through probabilistic forecasts is also of value as
it enables planners to manage associated risks. In this paper, we
investigate the application of hierarchical forecasting methods for
producing a probabilistic forecast of daily incident attended up to 84
days ahead, using different forecasting methods.

Our results indicates that forecast reconciliation can not only
contribute to a more coordinated efforts through producing coherent
forecasts when it comes to using forecasting to inform decisions in EMS,
but also it can result in forecast accuracy improvement. Our proposed
forecasting models combined with reconciliation approaches outperform
the empirical distribution benchmark. We show that a substantial
forecast improvement can be achieved at higher levels of hierarchical
and grouped structure time series by applying forecast reconciliation
methods. When a point forecast is of interest at the bottom level of
series, we observe that reconciliation may not improve the forecast
accuracy, if the bottom series are noisy and lack systematic patterns.
However, forecast reconciliation may result in more forecast results for
bottom series, if we are interested in the tails of forecast
distribution rather than just center measures like mean (i.e.~point
forecast). Producing consistent forecast are also crucial for informing
planning activities, we also demonstrate that proposed models produce
consistent forecasts across all forecast horizons. Therefore, we
recommend that forecast recociliation approaches to be adopted for
routine use in EMS, whenever hierarchical and/or grouped time series
data need to be forecasted. Moreover, we found that using an ensemble
forecasting model, combining al models developed in this paper, instead
of using each individualy, works remarkably well for our mixed
hierarchical \& grouped stucture.

Further research could investigate the practical benefits of
probabilistic hierarchical forecasting in EMS. Linking forecasts with
its utilities (response time, resource utilisation, etc) can offer an
opportunity to maximise benefits through more holistic planning
approach. While, we generated count attend incident for the base
forecast usign Poisson regression models, however the reconciles
forecast is snot yet count. This could be also an avenue for further
research.
